Enabling inference for context-dependent models of mutation by bounding the propagation of dependency

### 2022-03-07 writeups/context-dependence.tex
Will comments

    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
    which acts effectively as a selective pressure against G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
    If the mutation occurs at rate $\mu$ per individual,
    the total rate it occurs at in the population is $\mu N_e$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
    This is incorporated as follows: set the energy of patterns \nC{} and \nG{} to be $s$,
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
    with advantage $e$ relative to the per-individual, per-site mutation rate:
    \begin{align*}
        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
    \end{align*}
    With these definitions,
    if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation
    appears and successfully takes over in the population.

@@

\begin{example}[Gibbs sampling of the Ising model]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as a string of $L$ magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{L-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^L x_i$.
    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following process, also known as ``Glauber dynamics'', preserves the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+\exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$
        \end{tabular}
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular}
    \end{center}
    and
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}
\end{example}

\paragraph{The generator matrix}
We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^L \times |\calS|^L$ matrix $G(L)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^L$ jumps to state $y \in \calS^L$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le L$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.
\revpoint{2}{3}

@@

The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^L$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^L \times |\calS|^L$ matrix $G(L)$
is clearly infeasible even for moderately sized $L$.
We can, however, compute \eqref{eqn:full_likelihood} with smaller $G(n)$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, this ignores dependencies between neighboring blocks,
so it is not clear how good this approximation might be.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

@@

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that will be sufficient to compute the full likelihood,
at least arbitrarily.
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
(see Figure \ref{fig:Tmers}).
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
We emphasize that the overhang of the T-mers is typically bigger than the
context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).

These $p_{n,\ell,r}(x,y)$ probabilities can be found by marginalizing $p_n$,
summing over the possible initial and final substrings of $x$
(denoted $a$ and $b$):
\begin{align}
    p_{n,\ell,r}(t;x,y)
    &=
    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) ,
    \quad
    \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
    \label{eq:pmMarg}
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
the probability a given subsequence of length $n$ matching $x$
is later observed to have its center matching $y$
is close to $p_{n,\ell,r}(x,y)$, regardless of wider context.
Formally, for each $i$,
\begin{align} \label{eqn:window_approx}
    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
    \approx
    p_{n,\ell,r}(x,y)
    ,
\end{align}
with the approximation getting better for longer $\ell$ and $r$.

Equation \eqref{eq:pmMarg} could then be used to compute a composite likelihood,
which we denote $A_{n, \ell, r}(x,y)$.
This is the product of the probabilities
of transition for all $(n, \ell, r)$ T-mers between $x$ to $y$, i.e.,
with $m = n - \ell - r$,
\begin{align}\label{eqn:composite_like}
    A_{n, \ell, r}(x,y)
        =
        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(x_i^{(n)}, y_{i+\ell}^{(m)}\right).
\end{align}
To keep notation simple we've ignored the boundaries of the sequence;
to deal properly with these,
on the right-hand side of the equation $\ell$ should be replaced with $\min(\ell, \max(i-1, 0))$,
and $r$ should be replaced with $\min(r, \max(n-i-\ell-m+1, 0))$.

@@

Here is a heuristic argument, depicted in Figure~\ref{fig:Tmers}C.
Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
with $\ell = r$.
Then we know that site $i$ approximately depends only on mutations within distance $\ell$.
Furthermore, the sites up to distance $2 \ell$ to the left of $i$
also depend on these mutations.
We can now use \eqref{eqn:window_approx} to restrict the dependency on $x$.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
This suggests that the following approximation should be good:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell+1,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }.
    \end{split}
\end{align}

Since the full likelihood is a product across sites,
we can then simply express an approximate full likelihood as
a ratio of two composite likelihoods:
\[
    \like(y \mid x) \approx \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
\]
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}

@@

For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
for various $t$ (and $m = n - \ell - r$).
This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
\begin{align}
    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
\end{align}
where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long (i.e., length $n$) patterns $c$ that match the shorter (i.e., length $m$) pattern $b$
as in \eqref{eq:pmMarg}:
if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^{(m)}=b$ and $U_{cb}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{n,\ell,r}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(n)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

@@

Instead, we will compute the probability that
we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$,
summing over possible (long) patterns at the root.
Concretely: pick a random location $I$ in the sequence
and let $X_{I-\ell}^{(n)}(0) = \rho_I$ be the (long) pattern seen there at the root.
Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho_I=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho_I$ is proportional to
% \[
%     \P\{ \rho_I=z \mid X_{I-\ell}^{(n)}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(m)}(t_Y)=y \}
    =
    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
\end{align}
Denote this quantity $p_{n,\ell,r}(x, y)$.
This can be computed without much more effort than the simpler case above,
given the frequencies at the root, as described in greater generality below.
These calculations can be done using the same sparse matrix methods as above.
For instance, to compute
\[
    \sum_{z \in \calS^{n}}
        \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
    =
    \sum_{z \in \calS^{n}}
    \left( e^{t_X G(n)} \right)_{zx}
    \pi(z)
    \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_Y G(n)} \right)_{zw} U_{wy}$ as above,
multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $e^{t_X G(n)}$.

@@

\subsection{Acknowledgements}
The initial impetus for this project came from Matt Dean;
along the way substantial encouragement and useful suggestions were provided by
Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
Many thanks are also due to Jessica Crisci for curating the human-chimp data.
The majority of this work was done while PR was at USC.

>>>>

    Counteracting this trend towards decreased G/C
    is GC-biased gene conversion \citep{glemin2015quantification},
    which acts effectively as a selective pressure in favor of G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
    If the mutation occurs at rate $\mu$ per individual,
    the total rate it occurs at in the population is $\mu N_e$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected for.
    This is incorporated as follows: set the energy of patterns \nC{} and \nG{} to be $s$,
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    We then expect $s$ to be positive.
    Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
    with advantage $e$ relative to the per-individual, per-site mutation rate:
    \begin{align*}
        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
    \end{align*}
    With these definitions,
    if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation
    appears and successfully takes over in the population.

@@

\begin{example}[Gibbs sampling of the Ising model]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as a string of $L$ magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_{i=1}^{L-1} x_i x_{i+1} - \frac{1}{2} \gamma \sum_{i=1}^L x_i$.
    Here $\beta$ controls the strength of coupling between neighboring dipoles,
    and $\gamma$ controls the strength of an external magnetic field
    (and both are scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following process, also known as ``Glauber dynamics'', preserves the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+\exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$
        \end{tabular}
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular}
    \end{center}
    and
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}
\end{example}

\paragraph{The generator matrix}
We now describe concretely how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^L \times |\calS|^L$ matrix $G(L)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^L$ jumps to state $y \in \calS^L$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le L$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+j} = u_j \quad &\text{for } 0 \le j < h \\
  y_{i+j} = v_j \quad &\text{for } 0 \le j < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e., if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.
\revpoint{2}{3}

@@

The problem at hand is to infer the parameters of the model based on the final state of the model after it has evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^L$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^L \times |\calS|^L$ matrix $G(L)$
is clearly infeasible even for moderately sized $L$.
We can, however, compute \eqref{eqn:full_likelihood} with smaller $G(n)$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(t;x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, this ignores dependencies between neighboring blocks,
so it is not clear how good this approximation might be.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

@@

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that will be sufficient to compute the full likelihood,
at least approximately.
To do this, let $p_{n,\ell,r}(t;x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
(see Figure \ref{fig:Tmers}).
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
We emphasize that the overhang of the T-mers is typically bigger than the
context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).

These $p_{n,\ell,r}(t;x,y)$ probabilities can be found by marginalizing $p_n$,
summing over the possible initial and final substrings of $x$
(denoted $a$ and $b$):
\begin{align}
    p_{n,\ell,r}(t;x,y)
    &=
    \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) ,
    \quad
    \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
    \label{eq:pmMarg}
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,r}(t;x,y)$ to be asymptotically correct when $\ell$ and $r$ are large (and $t$ is fixed).
In Appendix~\ref{ss:approx_pf}, we show that this is so, namely that
the probability a given subsequence of length $n$ matching $x$
is later observed to have its center matching $y$
is close to $p_{n,\ell,r}(t;x,y)$, regardless of wider context.
Formally, for each $i$,
\begin{align} \label{eqn:window_approx}
    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
    \approx
    p_{n,\ell,r}(t;x,y)
    ,
\end{align}
with the approximation getting better for longer $\ell$ and $r$.

Equation \eqref{eq:pmMarg} could then be used to compute a composite likelihood,
which we denote $A_{n, \ell, r}(x,y)$.
This is the product of the probabilities
of transition for all $(n, \ell, r)$ T-mers between $x$ to $y$, i.e.,
with $m = n - \ell - r$,
\begin{align}\label{eqn:composite_like}
    A_{n, \ell, r}(x,y)
        =
        \prod_{i=1}^{L-m+1} p_{n, \ell, r}\left(t;x_i^{(n)}, y_{i+\ell}^{(m)}\right).
\end{align}
To keep notation simple we've ignored the boundaries of the sequence;
to deal properly with these,
on the right-hand side of the equation $\ell$ should be replaced with $\min(\ell, \max(i-1, 0))$,
and $r$ should be replaced with $\min(r, \max(n-i-\ell-m+1, 0))$.

@@

Here is a heuristic argument, depicted in Figure~\ref{fig:Tmers}C.
Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
with $\ell = r$.
Then we know that site $i$ approximately depends only on mutations within distance $\ell$.
Furthermore, the sites up to distance $2 \ell$ to the left of $i$
also depend on these mutations.
We can now use \eqref{eqn:window_approx} to restrict the dependency on $x$.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
This suggests that the following approximation should be good:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{4\ell+1,\ell,\ell}(t; x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{4\ell+1,\ell,\ell+1}(t; x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }.
    \end{split}
\end{align}
\revpoint{2}{6}

Since the full likelihood is a product across sites,
we can then simply express an approximate full likelihood as
a ratio of two composite likelihoods:
\[
    \like(y \mid x) \approx \frac{ A_{4\ell+1, \ell, \ell}(x,y) }{ A_{4\ell+1, \ell, \ell+1}(x,y) }.
\]
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(t;a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(t;c,d)
% \end{align}

@@

For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
for various $t$ (and $m = n - \ell - r$).
This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
\begin{align}
    p_{n}(t;a,c) = \left( e^{t G(n)} \right)_{a,c} ,
\end{align}
where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long (i.e., length $n$) patterns $c$ that match the shorter (i.e., length $m$) pattern $b$
as in \eqref{eq:pmMarg}:
if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^{(m)}=b$ and $U_{cb}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{n,\ell,r}(t;a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(n)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

@@

Instead, we will compute the probability that
we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$,
summing over possible (long) patterns at the root.
Concretely: pick a random location $I$ in the sequence
and let $X_{I-\ell}^{(n)}(0) = \rho_I$ be the (long) pattern seen there at the root.
Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho_I=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho_I$ is proportional to
% \[
%     \P\{ \rho_I=z \mid X_{I-\ell}^{(n)}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(m)}(t_Y)=y \}
    =
    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X;z,x) \, p_{n,\ell,r}(t_Y;z,y) .
\end{align}
Denote this quantity $p_{n,\ell,r}(x, y)$.
This can be computed without much more effort than the simpler case above,
given the frequencies at the root, as described in greater generality below.
These calculations can be done using the same sparse matrix methods as above.
For instance, to compute
\[
    \sum_{z \in \calS^{n}}
        \pi(z) \, p_{n,0,0}(t_X;z,x) \, p_{n,\ell,r}(t_Y;z,y)
    =
    \sum_{z \in \calS^{n}}
    \left( e^{t_X G(n)} \right)_{zx}
    \pi(z)
    \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_Y G(n)} \right)_{zw} U_{wy}$ as above,
multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $e^{t_X G(n)}$.

@@

\subsection{Acknowledgements}
The initial impetus for this project came from Matt Dean;
along the way substantial encouragement and useful suggestions were provided by
Will DeWitt, Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
Many thanks are also due to Jessica Crisci for curating the human-chimp data.
The majority of this work was done while PR was at USC.

### 2022-03-07 writeups/reviews/reviews.v1.tex
Will comments

\begin{point}{3d}
There is an inconsistency between equations (5), (6), and the
equation right below (6). In the numerator the identification between
the symbols must be that (n,l,r) in (5) is (2l+1,l+1,l) in (6).
However, the upper index on x is n in (5) but 4l+1 != 2l+1 in (6).
That is true in the denominator as well. Also,
m=n-l-r=(2l+1)-(l+1)-l=0 != 2l+1 (if the authors meant that n was
actually 4l+1, 4l+1-(l+1)-l=2l != 2l+1 does not work much better
either) and in the denominator it is not clear why m should change
given that l and r just get swapped and m only depends on their sum.
Maybe this referee totally misunderstands the notation here, but this
fact alone means that something needs to be improved in the
presentation in this area.
\end{point}

\reply{
TODO: Sorry about this. I suggest the solution is to change the
first subscript on p to 4l+1.
}

>>>>

\begin{point}{}
There is an inconsistency between equations \eqref{eqn:composite_like}, \eqref{eqn:full_approx},
and the equation right below \eqref{eqn:full_approx}. In the numerator the identification between
the symbols must be that $(n,\ell,r)$
in \eqref{eqn:composite_like} is $(2\ell+1,\ell+1, \ell)$ in \eqref{eqn:full_approx}.
However, the upper index on $x$ is $n$ in \eqref{eqn:composite_like}
but $4\ell+1 \neq 2\ell+1$ in \eqref{eqn:full_approx}.
That is true in the denominator as well. Also,
$m=n-\ell-r=(2\ell+1)-(\ell+1)-\ell=0 \neq 2\ell+1$
(if the authors meant that $n$ was
actually $4\ell+1$, then $4\ell+1-(\ell+1)-\ell=2\ell != 2\ell+1$ does not work much better
either) and in the denominator it is not clear why $m$ should change
given that $\ell$ and $r$ just get swapped and m only depends on their sum.
Maybe this referee totally misunderstands the notation here, but this
fact alone means that something needs to be improved in the
presentation in this area.
\end{point}

\reply{
Whoops - sorry about that.
The inconsistency stemmed from our switching around notation in a previous draft
but forgetting to update it in equation \eqref{eqn:full_approx}.
(This is confusing, which is why we have Figure~\ref{fig:Tmers}C to refer to.)
The notation is now fixed. \revref
}
